---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data  
This part of the code attaches the dplyr library, reads the data from the .csv file and saves it as the dataframe. For the sake of the futher analysis I change the format of the "date" variable to the POSIXct date format.  

```{r}
library(dplyr)
ActivityData<-read.csv(file="activity.csv", header=TRUE, sep=",")
ActivityData$date<-as.POSIXct(ActivityData$date,format='%Y-%m-%d')
summary(ActivityData)
```
## What is mean total number of steps taken per day?
This part of the code creates the dataframe that contains the total and average number of steps taken per day and creates the histogram of the total number of steps. Two horizontal lines define the mean and the median of the total number of steps taken per day. The corresponding values of the mean and the median total number of steps are given after the histogram together with the code generated these statistics.
```{r}

# Calcuate the total number of steps taken each day
DailyActivity<-group_by(ActivityData,ActivityData$date)
DailySummary<-summarize(DailyActivity,TotalSteps=sum(steps,na.rm=TRUE),AverageSteps=mean(steps,na.rm=TRUE))
# Create the histogram of the total steps
hist(DailySummary$TotalSteps,
     col="peachpuff",
     border="black",
     xlab="Number of steps",
     main="Total number of steps taken per day")
abline(v=mean(DailySummary$TotalSteps),
       col="royalblue",
       lwd=2)
abline(v=median(DailySummary$TotalSteps),
       col="red",
       lwd=2)
legend(x="topright",
       c("Mean","Median"),
       col=c("royalblue","red"),
       lwd=c(2,2))
## copy the histogram to a png file
dev.copy(png,file="figure\histogram.png",width=480,height=480)

## close the png device
dev.off()
```

Mean total number of steps:
```{r}
mean(DailySummary$TotalSteps)
```
Median total number of steps:
```{r}
median(DailySummary$TotalSteps)
```
## What is the average daily activity pattern?

This part of the code creates the dataframe with the average number of steps taken within particular 5-minute interval, what allows to check average intradaily activity of a person. 

```{r}

# Calcuate the average number of steps taken within 5-minute intervals
IntradailyActivity<-group_by(ActivityData,ActivityData$interval)
IntradailySummary<-summarize(IntradailyActivity,TotalSteps=sum(steps,na.rm=TRUE),AverageSteps=mean(steps,na.rm=TRUE))
# Find the interval within which the maximum number of steps was made on average
IntervalMaxSteps<-filter(IntradailySummary,AverageSteps==max(AverageSteps))

# Create the time series plot of the average number of steps
plot(IntradailySummary$`ActivityData$interval`,IntradailySummary$AverageSteps,type='l',ylab="Number of steps",xlab="Interval")
# add the vertical line to show the interval with the maximum averaged number of steps
abline(v=IntervalMaxSteps[1],
       col="royalblue",
       lwd=2)

## copy the histogram to a png file
dev.copy(png,file="figure\TimeSeriesPlot.png",width=480,height=480)

## close the png device
dev.off()
```

The indicator of the 5-minute interval that, on average across all the days in the dataset, contains the maximum number of steps:
```{r}
IntervalMaxSteps
```
## Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)


Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Create a new dataset that is equal to the original dataset but with the missing data filled in.

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
