---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

This part of the code attaches the dplyr library, reads the data from the .csv file and saves it as the dataframe. For the sake of the futher analysis I change the format of the "date" variable to the POSIXct date format.  

```{r}
library(dplyr)
ActivityData<-read.csv(file="activity.csv", header=TRUE, sep=",")
ActivityData$date<-as.POSIXct(ActivityData$date,format='%Y-%m-%d')
```

## What is mean total number of steps taken per day?

```{r}

# Calcuate the total number of steps taken each day
DailyActivity<-group_by(ActivityData,ActivityData$date)
DailySummary<-summarize(DailyActivity,steps=sum(steps,na.rm=TRUE),interval=mean(interval,na.rm=TRUE))
# Create the histogram
hist(DailySummary$steps)
```



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
